---
title: "Querying .rete graphs from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Querying .rete graphs from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A `.rete` file is a single, immutable RDF graph image designed to be queried
**in place**: locally with positional reads, or straight off any HTTP host
that honors `Range` requests — S3, R2, GitHub releases, a plain web server.
Only the byte ranges a query touches are fetched, so multi-gigabyte graphs
on static storage stay interactive with no database server.

This vignette is self-contained: it builds a small graph in memory and
queries it. The remote examples at the end are shown as-is (they need the
network) — they behave identically.

## Build a graph, query it

`rete_build()` assembles a complete `.rete` file image — dictionary, triple
indexes, community pyramid, and optionally an embedded Dataset Card — from
RDF text:

```{r}
library(rete)

nt <- paste(
  '<urn:x:alice> <urn:rdf:type> <urn:x:Person> .',
  '<urn:x:alice> <urn:x:name> "Alice" .',
  '<urn:x:alice> <urn:x:age> "42"^^<http://www.w3.org/2001/XMLSchema#integer> .',
  '<urn:x:bob> <urn:rdf:type> <urn:x:Person> .',
  '<urn:x:bob> <urn:x:name> "Bob" .',
  '<urn:x:bob> <urn:x:knows> <urn:x:alice> .',
  sep = "\n"
)

img <- rete_build(nt, card = list(
  title = "Tiny people graph",
  description = "Six triples, for the vignette",
  license = "CC0-1.0"
))
g <- rete_open(img)
g
```

`rete_query()` returns what an R user expects — SELECT gives a `data.frame`
with one column per variable, IRI brackets stripped and typed literals
coerced (integers, doubles, logicals):

```{r}
rete_query(g, "SELECT ?name ?age WHERE {
  ?p <urn:x:name> ?name .
  OPTIONAL { ?p <urn:x:age> ?age }
} ORDER BY ?name")
```

ASK returns a logical scalar; CONSTRUCT and DESCRIBE return a
subject/predicate/object frame:

```{r}
rete_query(g, "ASK { ?s <urn:x:knows> ?o }")
rete_query(g, "CONSTRUCT { ?o <urn:x:knownBy> ?s } WHERE { ?s <urn:x:knows> ?o }")
```

`rete_query_raw()` keeps full N-Triples token fidelity when the coercions
are too helpful.

## Files describe themselves

A `.rete` can carry its own Dataset Card — title, license, provenance, and
runnable example queries — so a file you have never seen explains itself:

```{r}
card <- rete_card(g)
card$title
card$quad_count

rete_schema(g)$classes
```

Write the image to disk with `writeBin(img, "people.rete")`; local files
open lazily too:

```{r}
path <- tempfile(fileext = ".rete")
writeBin(as.vector(img), path)
rete_info(rete_open(path))
```

## Remote graphs, lazily

The same calls work against a URL — the host only needs to answer `Range`
requests with `206 Partial Content`, which any object store or CDN does:

```{r, eval = FALSE}
g <- rete_open("https://data.graphplaza.com/boe/boe.rete")
rete_query(g, "SELECT ?s ?p ?o WHERE { ?s ?p ?o } LIMIT 3")

rete_stats(g)
#> $fileLength
#> [1] 6958628
#> $bytes
#> [1] 1703936      # ~24% of the file, cached for the next query
#> $requests
#> [1] 13
```

`rete_stats()` is the honesty meter: it reports the bytes physically
fetched since open. Repeated queries reuse the decoded dictionary and index
tiles on the handle, so they get faster and cheaper.

Files that embed example queries hand you a starting point:

```{r, eval = FALSE}
ex <- rete_examples(g)
ex[, c("title", "question")]
rete_query(g, ex$sparql[[1]])
```

And files whose ontology supports it answer with OWL 2 QL entailment,
computed by query rewriting — no materialization, works remotely:

```{r, eval = FALSE}
rete_query(g, query, reason = TRUE)
```

## Where to go next

- The [R API guide](https://caviri.github.io/rete/r.html) covers every
  function; `?rete_open` and friends are the reference pages.
- The [format documentation](https://caviri.github.io/rete/) explains what
  a `.rete` is, how to build big ones with the CLI, and how to host them.
- The [browser playground](https://caviri.github.io/rete/playground.html)
  queries the same files with zero install — handy for sharing results.
